---
title: "Practice project: Admissions2 data"
author: "Joshua Megnauth"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    df_print: paged
---

## Introduction
**THIS PROJECT IS CURRENTLY INCOMPLETE.**

```{r load_libraries_data, warning=FALSE, message=FALSE}
library(kableExtra)
library(parallel)
library(doParallel)
library(pROC)
library(ranger)
library(caret)
library(tidyverse)
library(ggplot2)
library(broom)

# Dracula colors
# https://draculatheme.com/
# Obviously not following the theme yet.
PINK <- "#ff79c6"
CYAN <- "#8be9fd"
GREEN <- "#50fa7b"
PURPLE <- "#bd93f9"
RED <- "#ff5555"
BACKGROUND <- "#282a36"

# Threshold for the outcomes (i.e. above THRESHOLD = a success)
THRESHOLD <- 0.65

admissions_name <- "Admission_Predict_Ver1.1.csv"
admissions_cols <- c("id", "gre", "toefl", "uni_ratings", "statement",
                     "letter", "cgpa", "research", "prob_admit")
admissions_types <- "iiifdddld"

admissions <- 
  read_csv(admissions_name, col_names = admissions_cols,
           col_types = admissions_types, skip = 1) %>%
  select(-id) %>%
  mutate(research = as_factor(if_else(research, "Yes", "No")))

cluster_mp <- makeCluster(detectCores() - 1)
registerDoParallel(cluster_mp)

options(scipen = 16)
set.seed(314)
rm(admissions_name, admissions_cols, admissions_types)
```

The [graduate admissions data](https://www.kaggle.com/mohansacharya/graduate-admissions) data set is not immediately attractive at first glance. The data set is sparse, utilitarian, and tractable. Utility is a function of the data's admittedly humdrum goal of predicting graduate school admissions from dreary tests. However, the data set is also yielding due to the directness of the variables and the lack of cleaning required.

For those reasons I had been drawn to this data as a first, small project to play around with in R and Python.

Let's quickly go over the code above. **THRESHOLD** is a constant to separate the positive predictions from the negative---that is, admissions from non-admissions. I selected **0.65** as a probability after exploring the data below which means that the value is a judgement call rather than a science. As a parameter, a statistician or data scientist may tune the threshold for separating the classes depending on the teleology of the problem. Finally, 0.65 is hardly perfect, but I found the value to perform well.

Anyway, the code to load the data isn't all that interesting. I manually skipped the header row, renamed the columns, and set the types. I converted _research_ to a factor with the levels renamed as "Yes" and "No" as TRUE/FALSE are invalid names in R. [Caret](https://topepo.github.io/caret/)---a machine learning library developed by Max Kuhn---throws an exception when invalid names are used.

I'm also putting my [i7](https://en.wikipedia.org/wiki/List_of_Intel_Core_i7_microprocessors) to work with multiprocessing via [doParallel](https://cran.r-project.org/web/packages/doParallel/index.html). Training hundreds of models is an example of a problem that is [embarrassingly parallel](https://en.wikipedia.org/wiki/Embarrassingly_parallel), so I suggest that anyone using cross validation and grid search look into multiprocessing via Python or R.

As of 8/2/2020 I have set a seed of **314.** My random seed will always be 324, 314, 428, or 42. My code is open source with each commit available on GitHub to confirm I am not tampering with random seeds.

## Exploring the data
```{r tests_plots}
standardize <- function(variable) {
  (variable - mean(variable))/sd(variable)
}

thresh_to_fac <- function(data, threshold = THRESHOLD) {
  # Factor levels are created by order, so I need to ensure the order for
  # the negative class is explicit.
  temp <- as_factor(if_else(data > threshold, "Yes", "No"))
  relevel(temp, "No")
}

train_test_split <- function(X, y, test_size = .2) {
  len <- length(X)
  rows <- sample(len)
  split <- test_size * len
  
  X_test <- X[1:split]
  y_test <- y[1:split]

  X_train <- X[split + 1:len]
  y_train <- y[split + 1:len]
  
  list(X_train, y_train, X_test, y_test)
}

admissions %>%
  mutate(gre_stdize = standardize(gre),
         toefl_stdize = standardize(toefl)) %>%
  ggplot() +
  geom_histogram(aes(gre_stdize), binwidth = .25,
                 color = BACKGROUND, fill = GREEN) +
  geom_histogram(aes(toefl_stdize), binwidth = .25,
                 color = BACKGROUND, fill = PINK, alpha = 0.5) +
  xlab("GRE [green] and TOEFL [pink]") +
  ylab("Frequency") +
  ggtitle("Standardized GRE and TOEFL scores")
```

I standardized the variables for GRE and TOEFL scores due to the different scales. Working with standardized or normalized variables is often just plain easier. The process of standardizing a variable is simple. We'd like a standard deviation of 1 and a mean of zero, so we can simple subtract the mean of the variable from each score and divide by the standard deviation. In other words, we're taking the z score of each value.

$$z = \frac{x - \mu}{\sigma}$$

The distributions for GRE and TOEFL scores seem fairly similar as we can see from the histograms. In some cases, such as the range from a standard deviation of **0 to 1**, the distributions are near identical.

I wrote a function, **thresh_to_fac()**, to encapsulate converting probabilities to a factor using a threshold. The function cleans up my code a bit when I have to convert probabilities. My function first converts all cases where a column (the parameter **data**) is greater than the threshold to "Yes" or "No." The "Yes" and "No" values are converted to a factor which in turn is releveled to ensure "No" is the first or zeroth case.

```{r probability_admission}
admissions %>%
  ggplot() +
  geom_density(aes(prob_admit, fill = research), alpha = 0.5) +
  scale_fill_manual(values = c(PINK, CYAN)) +
  labs(fill = "Research") +
  xlab("Admission") +
  ylab("Probability") +
  ggtitle("Probability of admission by research experience")
```

The probability of admission seems normally distributed for students without research experience while positively skewed for students with projects. Both modes peak past the midway point of 0.5. My initial guess is that setting a threshold of 0.65 would catch more _true positives_ than using 0.5. I wouldn't happily cartwheel around my room at 50-50 odds for school admissions. Likewise, a strict separation at 50% could risk incorrect predictions especially when reasonable students wouldn't bet on such low odds.

A higher threshold logically risks more false negatives---cases where students would be admitted but the model predicted rejection. However, a greater false negatives rate is less risky than false positives here in my opinion. Applying to schools only to be declined seems more individually taxing than false negatives unless the rate for the latter is unreasonably high.

```{r}
admissions %>%
  select(gre, toefl, statement, letter) %>%
  rename(GRE = gre, TOEFL = toefl, SoP = statement,
         LoR = letter) %>%
  cor() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

GRE, TOEFL, Statements of Purpose ratings, and Letters of Recommendation ratings are all at least somewhat correlated. The two tests, GRE and TOEFL, are greatly correlated with $R = 0.83$. Ditching at least one of these variables may improve modeling or, at the very least, prevent wonky coefficients.

## Logistic regression

```{r naive_glm}
admissions <- admissions %>%
  mutate(y_admit = thresh_to_fac(prob_admit))

basic_form <- y_admit ~ gre + uni_ratings + cgpa + statement + letter + research
basic_mod <- glm(basic_form, "binomial", admissions)

tidy(basic_mod, conf.int = TRUE, exponentiate = TRUE)
```

My first model is conservative and straightforward. I included only one of the tests as the two test scores are correlated. Of course, a person/observation that has a certain score on one test is likely to have a similar grade on the other test. Likewise, G.P.A. shouldn't deviate too wildly from TOEFL or GRE scores, but I included the variable in this preliminary model to be safe.

The coefficients for the odds fall into the confidence intervals. The confidence intervals are a bit wonky. For example, the confidence intervals for _university ratings_ are rather wide. P-values measure the probability of results at least as extreme as what was observed given that the null hypothesis is true. In other words, the p-value is concerned with only the probability of extreme events in terms of the _data_ along with assuming the null hypothesis of no effect; we're not considering external information, and the model says nothing about truthiness. Only the odds coefficient of GRE scores, cumulative G.P.A., and letters of recommendation scores may not be due to random chance given $\alpha = .05$.

Now let's take a look at the appropriate coefficients! Each additional point gained on the GRE increases an applicant's acceptance odds by about 1.07. Cumulative G.P.A. 

The intercept is so low that it's basically zero. But, technically each applicant starts with an odds of .0000000000000000000000407829 if they managed to have a zero GPA, failed the GRE with a zero, and wrote random letters for their statement of purpose, et cetera...

### Diagnostics---logistic regression (unscaled)

Let's test predictive ability. I fit my basic model on all of the rows of data which in turn will be reused for prediction. Of course, I realize that precluding a hold out set as well as using training data for prediction are anti-patterns, but the model above is basic and experimentation is fun!

```{r basic_model_test, results="asis"}
confusion_wrapper <- function(y_true, y_pred) {
  # Create a confusion matrix object via caret
  confusion <- confusionMatrix(y_pred, reference = y_true, positive = "Yes")
  
  # Generate a nice table from the matrix via kableExtra
  cmatrix <-
    confusion$table %>%
    kable(caption = "Rows = Predictions") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
  
  # Likewise for the metrics
  metrics <-
    tidy(confusion) %>%
    select(term, estimate) %>%
    kable(col.names = c("Classification Function", "Value")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
  
  # Return both tables so they're automatically displayed
  list(ConfusionMatrix = cmatrix, Metrics = metrics)
}

basic_preds <- thresh_to_fac(predict(basic_mod, type = "response"))
confusion_wrapper(admissions$y_admit, basic_preds)
```

My basic model performs _pretty good_ considering I conservatively hand picked variables. With that said, I'll repeat that I _tested_ on my _training_ data so the model may be over fit with only decent performance. Total accuracy is the ratio of correctly separated true positive and true negatives over all of the classes total. Accuracy's major problem is that the two classes may be unbalanced. The canonical example considers the test for an affliction where most test results are negative. Let's say that 99% of the cases in a data set for some condition are negative. A classifier may achieve an accuracy score of 99% while predicting **literally every positive case incorrectly.**

The logistic model above achieves an accuracy of **86%** with a sensitivity (true positive rate) of **85.4%**. Specificity, or correctly sorting false cases, is similar at **87.4%**. The confusion matrix above shows **20 false positives** and **50 false negatives**. I set a threshold of 0.65 as discussed earlier. However, that high threshold may have borked the sorting ability of the model considering that the data is skewed towards positive cases as you can see below.

```{r cases_thresh_sixtyfive}
table(admissions$y_admit) %>%
  kable(col.names = c("Class", "Frequency")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

How about we check model performance with the threshold set at 0.5? For fun (and profit)!

```{r log_thresh_fifty, results="asis"}
# This code is dreadful. Sorry.
fifty_pred <- admissions %>%
  mutate(y_admit = thresh_to_fac(prob_admit, threshold = .5)) %>%
  glm(basic_form, "binomial", data = .) %>%
  predict(type = "response") %>%
  thresh_to_fac(threshold = .5)

fifty_true <- thresh_to_fac(admissions$prob_admit, .5)
confusion_wrapper(fifty_true, fifty_pred)

rm(fifty_pred, fifty_true)
```

Using my handy dandy ~~notebook~~ _thresh_to_fac()_ function, I generated a set of predictions as factors where the positive class is defined as **0.51** and up.

A threshold of 0.5 is ostensibly better as the model above sports an impressive **95%** accuracy with a sensitivity of **98.5%**. Specificity is a major downgrade at **48%.** Take a gander at the confusion matrix. My initial intuition from data exploration has been vindicated! Mostly! A threshold of 0.5 leads to very few total amount of "No" values. The 0.5 threshold model correctly predicts 454 of the positive class, but that metric is meaningless since most values are "Yes." A threshold of 0.5 under performs because the classes are disingenuously split even though the actual model technically performs well. Also, I'll reiterate that this model is fit on all of the data. The model is likely over fit both due to the data used as well as the split used.

I'll keep 0.65 as a threshold as the performance is stable.

```{r drop_prob_admit}
admissions <- admissions %>%
  select(-prob_admit)
```

## Bootstrap aggregated random forests [Baggin' CARTs]

Random forests is a clever name for a clever algorithm. An individual decision tree continually splits data by attempting to satisfy a criterion, such as maximizing information gain. The first split is known as the root node which corresponds to the primary, major schism determined by the algorithm. Each branch continues along the same process until the data is exhaustively categorized or a predefined stopping point is reached.

Of course, decision trees have more going in than a scant summary can cover, but that's the gist of the process. An individual decision tree lacks robustness because of its singularity. A single decision tree may hone in an optimal split, but a single split may be limiting in comparison to having a range of choices. Random forests grows many trees which are suboptimally learned leading to high bias. These trees understand a subset (bootstrap sample) of the data as well as certain patterns that are not ideal. However, random forest coalesces these patterns via voting which leads to stronger models overall. Using multiple trees solves another potential decision tree problem: over fitting. A decision tree must usually be cut or trimmed so that it doesn't learn a data set's noise. Fitting multiple weak learners helps random forests generalize an understanding of the data rather than learning its nuances.

```{r bagging_rf}
rf_traincontrol <- trainControl(method = "repeatedcv",
                                number = 5,
                                repeats = 5,
                                search = "random",
                                savePredictions = TRUE,
                                classProbs = TRUE,
                                summaryFunction = twoClassSummary,
                                allowParallel = TRUE)

# mtry_default <- floor(sqrt(ncol(admissions) - 1))
rf_grid <- expand.grid(mtry = 1:ncol(admissions) - 1,
                       splitrule = c("gini", "extratrees"),
                       min.node.size = c(1, 3, 5, 10))

rf_model_nopreproc <- train(y_admit ~ .,
                            method = "ranger",
                            trControl = rf_traincontrol,
                            tuneGrid = rf_grid,
                            metric = "ROC",
                            data = admissions)
```

The two highest rated answers on a question about [preprocessing for random forests](https://stats.stackexchange.com/questions/172842/best-practices-with-data-wrangling-before-running-random-forest-predictions) on Stack Overflow conflict. Wonderful! Random forests should theoretically be indifferent toward outliers or unimportant variables. One answer describes the algorithm as fairly "off the shelf," meaning that one may deploy a CART without much wrangling. Another answer explains that the theoretical prowess is sometimes wonky; all data should be seasoned a bit before cooking.

Scaled variables shouldn't harm nor help information gain at each cut of the tree, but I'll try the model both ways. Less complex models or modeling processes should always be preferred, of course, even if preprocessing doesn't strictly harm anything.

The hyperparameter **mtry** refers to the number of variables that may be possibly selected at each cut of the tree (this explanation may be off). By [default mtry](https://cran.r-project.org/web/packages/ranger/ranger.pdf) is set to the floor of the square root of the number of variables which is ~2 in this case.

Cross validation refers to a class of processes for testing model stability. Cross validation splits the data into _k_ folds and uses _k-1_ folds for training with the extra set used for testing. I'm using **repeated cross validation** as the data are rather limited at `r nrow(admissions)` rows. I'm avoiding splitting the data into training and test sets for the first model due to that reason as well. I originally used **leave one out cross validation** which trains the model _N_ times (k == N). LOOCV is clearly computational expensive but provides very conservative estimates of model performance. However, I've found that LOOCV, CV, and repeated CV all offer same conclusions since the model is fairly stable. In other words, error doesn't vary wildly across folds. While I could stick with k fold cross validation, repeated CV seems to be idiomatic for a data set of this size.

### Diagnostics---Random forest (unscaled)
```{r rf_no_preproc_diag, results="asis"}
rf_model_nopreproc$bestTune
plot(rf_model_nopreproc)

rf_y_pred_nopre <- relevel(predict(rf_model_nopreproc), "No")
confusion_wrapper(admissions$y_admit, rf_y_pred_nopre)
```

The model with the highest Receiver Operating Characteristic (ROC) ratio was trained with an **mtry** of 7 using the **extremely randomized trees (Extra Trees)** algorithm with a minimum node size of **10**. Using Gini impurity seems to work best with an _mtry_ of two with this data. Extra trees works best with an _mtry_ of six and a minimal node size of ten with this data.

## Random forests with preprocessing

```{r bagging_preproc_rf, warning=FALSE}
train_index <- createDataPartition(admissions$y_admit, p = 0.9)$Resample1

X_train <- admissions %>%
  slice(train_index) %>%
  select(-y_admit)

y_train <- admissions %>%
  slice(train_index) %>%
  select(y_admit) %>%
  unlist()

X_test <- admissions %>%
  slice(-train_index) %>%
  select(-y_admit)

y_test <- admissions %>%
  slice(-train_index) %>%
  select(y_admit) %>%
  unlist()

rf_model_preproc <- train(x = X_train,
                          y = y_train,
                          method = "ranger",
                          preProcess = c("scale", "pca"),
                          trControl = rf_traincontrol,
                          tuneGrid = rf_grid,
                          metric = "ROC")
```
My second random forests model preprocesses the variables by scaling numeric features and determining principal components via PCA. I also use training and holdout sets (which I should have just done initially as I gained nothing but created extra work for myself by modelling using all of the data).

### Diagnostics---random forest (scaled)
```{r rf_preproc_diagnostics, results="asis"}
rf_model_preproc$bestTune
rf_ypred_preproc <- relevel(predict(rf_model_preproc, newdata = X_test), "No")
confusion_wrapper(y_test, rf_ypred_preproc)
```
Unsurprisingly, a model that is trained on one segment of the data but tested on unseen data performs a bit worse than that model that is trained and tested on the same data. I mentioned above that I gave myself extra work by not using a training set. I can't really compare these two models as they were trained and tested on different data. However, I _really_ want to compare the performance of the different models. I'll retrain the logistic and unpreprocessed models on the training set then compare all of the models.

With that said, the confusion matrix above shows a very low specificity. The low row count of the data set coupled with the fact that the test set lacks observations as well seems to throw off predictions a bit.
## Feature engineering and a model comparison
Before I compare the models above properly I'd like to try feature engineering to see if I can improve performance.

```{r feature_engineering}
admissions_fe <-
  admissions %>%
  mutate(tests = gre + toefl,
         tests = standardize(tests),
         extra_docs = statement + letter,
         extra_docs = standardize(extra_docs),
         cgpa = standardize(cgpa)) %>%
  select(tests, extra_docs, cgpa, uni_ratings, research, y_admit)

# Train-test split
Xy_train_fe <- admissions_fe %>%
  slice(train_index)

X_test_fe <- admissions_fe %>%
  slice(-train_index) %>%
  select(-y_admit)
```

Right! So I added up the test scores then standardized the new variable. TOEFL and GRE were correlated, but I'm assuming simply combining them is a good way to use both variables. I kneaded statement of purpose and letter of recommendation in the same way. My assumption, which I'll test below, is that I can erode whatever negative effects correlation may have while squeezing out some extra performance. I also assume that logistic regression may be affected more by the feature engineering.

### Model comparison
```{r roc_auc, warning=FALSE, message=FALSE}
levels = c("No", "Yes")

# Logistic
logi_standard <- glm(basic_form, "binomial", admissions[train_index, ])
logi_standard_pred <- predict(logi_standard, newdata = X_test,
                              type = "response")

# Random forest (unscaled)
rf_standard <- train(x = X_train,
                     y = y_train,
                     method = "ranger",
                     trControl = rf_traincontrol,
                     tuneGrid = rf_grid,
                     metric = "ROC")
rf_standard_pred <- predict(rf_standard, newdata = X_test, type = "prob")$Yes

# Random forest (scaled)
rf_preproc_pred <- predict(rf_model_preproc, newdata = X_test,
                           type = "prob")$Yes

# Logistic (feature engineering)
logi_fe <- glm(y_admit ~ ., family = "binomial", data = Xy_train_fe)
logi_fe_pred <- predict(logi_fe, newdata = X_test_fe, type = "response")

# Random forest (feature engineering)
rf_fe <- train(y_admit ~ .,
               data = Xy_train_fe,
               method = "ranger",
               trControl = rf_traincontrol,
               tuneGrid = rf_grid,
               metric = "ROC")
rf_fe_pred <- predict(rf_fe, newdata = X_test_fe, type = "prob")$Yes

# ROC calculation via pROC
roc_logistic <- roc(response = y_test, predictor = logi_standard_pred,
                    levels = levels)
roc_nopreproc_rf <- roc(response = y_test, predictor = rf_standard_pred,
                        levels = levels)
roc_preproc_rf <- roc(response = y_test, predictor = rf_preproc_pred,
                      levels = levels)
roc_log_fe <- roc(response = y_test, predictor = logi_fe_pred, levels = levels)
roc_rf_fe <- roc(response = y_test, predictor = rf_fe_pred, levels = levels)

# List of the above calculations to Purrr on
roc_list <- list(Logistic = roc_logistic, RF_Preprocessed = roc_preproc_rf,
                 RF_NoPreprocess = roc_nopreproc_rf, LogisticFE = roc_log_fe,
                 RF_FE = roc_rf_fe)

# List of labels for Purrr.
roc_names = list("Logistic (unscaled)",
                 "Random forest (unscaled)",
                 "Random forest (scaled)",
                 "Logistic (feature engineering)",
                 "Random forest (feature engineering)")

# Actual labels to pass to ggroc. See explanation below.
roc_labels <- 
  roc_list %>%
  map(auc) %>%
  map_dbl(1) %>%
  map2_chr(roc_names, ~paste(.y, "AUC:", .x))

# ROC curves
ggroc(roc_list) +
  xlab("Specificity") +
  ylab("Sensitivity") +
  ggtitle("ROC curves") +
  scale_color_discrete(labels = roc_labels)
```

The code above looks very verbose, but I'm not doing anything really extravagant here besides [Purrr](https://purrr.tidyverse.org/) (meow). First, I retrain the models using the same subset of data---that is, the same split of rows. Next, I calculate predictions which are output as probabilities. Finally, I calculate the ROC using pROC and plot all of the curves on the same graph.

The Purrr code creates a list of strings that concatenate the names of the models with the AUC value. I map the [auc](https://rdrr.io/cran/pROC/man/auc.html) function onto the list of ROC objects followed by pulling out the second element from the new list of auc objects using **map_dbl** from Purrr. Auc's second element is the actual AUC. Finally, I concatenate each string from the list of names with the pulled AUC values.

```{r stop_cluster}
stopCluster(cluster_mp)
```
