---
title: "Kaggle Admissions2 Data Test"
author: "Joshua Megnauth"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    df_print: paged
---

## Introduction
```{r load_libraries_data, warning=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)
library(ggplot2)
library(broom)

# Dracula colors
# https://draculatheme.com/
# Obviously not following the theme yet.
PINK <- "#ff79c6"
CYAN <- "#8be9fd"
GREEN <- "#50fa7b"
PURPLE <- "#bd93f9"
RED <- "#ff5555"
BACKGROUND <- "#282a36"

admissions_name <- "Admission_Predict_Ver1.1.csv"
admissions_cols <- c("id", "gre", "toefl", "uni_ratings", "statement",
                     "letter", "cgpa", "research", "prob_admit")
admissions_types <- "iiifdddld"

admissions <- 
  read_csv(admissions_name, col_names = admissions_cols,
           col_types = admissions_types, skip = 1)

options(scipen = 100)
rm(admissions_name, admissions_cols, admissions_types)
```

The [graduate admissions data](https://www.kaggle.com/mohansacharya/graduate-admissions) data set is not immediately attractive at first glance. The data set is sparse, utilitarian, and tractable. Utility is a function of the data's admittedly humdrum goal of predicting graduate school admissions from dreary tests. However, the data set is also yielding due to the directness of the variables and the lack of cleaning required.

For those reasons I had been drawn to this data as a first, small project to play around with in R and Python.

Let's quickly go over the code above. 

## Exploring the data
```{r tests_plots}
standardize <- function(variable) {
  (variable - mean(variable))/sd(variable)
}

admissions %>%
  mutate(gre_stdize = standardize(gre),
         toefl_stdize = standardize(toefl)) %>%
  ggplot() +
  geom_histogram(aes(gre_stdize), binwidth = .25,
                 color = BACKGROUND, fill = GREEN) +
  geom_histogram(aes(toefl_stdize), binwidth = .25,
                 color = BACKGROUND, fill = PINK, alpha = 0.5) +
  xlab("GRE and TOEFL (standardized)") +
  ylab("Frequency") +
  ggtitle("Standardized GRE and TOEFL scores")
```

$$z = \frac{x - \mu}{\sigma}$$

[to do]
1. add titles for girraphs
2. Check distribution of statements of purpose/letters of recommendation

```{r probability_admission}
admissions %>%
  ggplot() +
  geom_density(aes(prob_admit, fill = research), alpha = 0.5) +
  scale_fill_manual(values = c(PINK, CYAN)) +
  labs(fill = "Research") +
  xlab("Admission") +
  ylab("Probability") +
  ggtitle("Probability of admission by research experience")
```

[notes]: probability with no research looks normal while both are distributed beyond 50%

```{r}
```

## A basic model

```{r naive_glm}
admissions <- admissions %>%
  mutate(y_admit = prob_admit > .65) %>%
  select(-prob_admit, -id)

basic_mod <- glm(y_admit ~ gre + uni_ratings + cgpa + statement + research,
                 "binomial", admissions)

tidy(basic_mod, conf.int = TRUE, exponentiate = TRUE)
```
[notes]: straightforward model, no scaling, cgpa may be bad to include
### Diagnostics

```{r vif}

```
