---
title: "Kaggle Admissions2 Data Test"
author: "Joshua Megnauth"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    df_print: paged
---

## Introduction
```{r load_libraries_data, warning=FALSE, message=FALSE}
library(kableExtra)
library(caret)
library(tidyverse)
library(ggplot2)
library(broom)

# Dracula colors
# https://draculatheme.com/
# Obviously not following the theme yet.
PINK <- "#ff79c6"
CYAN <- "#8be9fd"
GREEN <- "#50fa7b"
PURPLE <- "#bd93f9"
RED <- "#ff5555"
BACKGROUND <- "#282a36"

# Threshold for the outcomes (i.e. above THRESHOLD = a success)
THRESHOLD <- 0.65

admissions_name <- "Admission_Predict_Ver1.1.csv"
admissions_cols <- c("id", "gre", "toefl", "uni_ratings", "statement",
                     "letter", "cgpa", "research", "prob_admit")
admissions_types <- "iiifdddld"

admissions <- 
  read_csv(admissions_name, col_names = admissions_cols,
           col_types = admissions_types, skip = 1) %>%
  select(-id)

options(scipen = 16)
rm(admissions_name, admissions_cols, admissions_types)
```

The [graduate admissions data](https://www.kaggle.com/mohansacharya/graduate-admissions) data set is not immediately attractive at first glance. The data set is sparse, utilitarian, and tractable. Utility is a function of the data's admittedly humdrum goal of predicting graduate school admissions from dreary tests. However, the data set is also yielding due to the directness of the variables and the lack of cleaning required.

For those reasons I had been drawn to this data as a first, small project to play around with in R and Python.

Let's quickly go over the code above. 

## Exploring the data
```{r tests_plots}
standardize <- function(variable) {
  (variable - mean(variable))/sd(variable)
}

admissions %>%
  mutate(gre_stdize = standardize(gre),
         toefl_stdize = standardize(toefl)) %>%
  ggplot() +
  geom_histogram(aes(gre_stdize), binwidth = .25,
                 color = BACKGROUND, fill = GREEN) +
  geom_histogram(aes(toefl_stdize), binwidth = .25,
                 color = BACKGROUND, fill = PINK, alpha = 0.5) +
  xlab("GRE and TOEFL (standardized)") +
  ylab("Frequency") +
  ggtitle("Standardized GRE and TOEFL scores")
```

$$z = \frac{x - \mu}{\sigma}$$

[to do]
Explain the standardization formula.
Change y to a factor since caret doesn't like logicals for some reason.

```{r probability_admission}
admissions %>%
  ggplot() +
  geom_density(aes(prob_admit, fill = research), alpha = 0.5) +
  scale_fill_manual(values = c(PINK, CYAN)) +
  labs(fill = "Research") +
  xlab("Admission") +
  ylab("Probability") +
  ggtitle("Probability of admission by research experience")
```

[notes]: probability with no research looks normal while both are distributed beyond 50%

```{r}
admissions %>%
  select(gre, toefl, statement, letter) %>%
  cor() %>%
  kable(col.names = c("GRE", "TOEFL", "SoP", "LoR")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

admissions %>%
  select(gre, toefl, statement, letter) %>%
  cor() %>%
  data.frame() %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## A basic model

```{r naive_glm}
admissions <- admissions %>%
  mutate(y_admit = prob_admit > THRESHOLD) %>%
  select(-prob_admit)

basic_form <- y_admit ~ gre + uni_ratings + cgpa + statement + letter + research
basic_mod <- glm(basic_form, "binomial", admissions)

tidy(basic_mod, conf.int = TRUE, exponentiate = TRUE)
```

My first model is conservative and straightforward. I included only one of the tests as the two test scores are correlated. Of course, a person/observation that has a certain score on one test is likely to have a similar grade on the other test. Likewise, G.P.A. shouldn't deviate too wildly from TOEFL or GRE scores, but I included the variable in this preliminary model to be safe.

The coefficients for the odds fall into the confidence intervals. The confidence intervals are a bit wonky. For example, the confidence intervals for _university ratings_ are rather wide. P-values measure the probability of results at least as extreme as what was observed given that the null hypothesis is true. Only GRE scores, cumulative G.P.A., and letters of recommendation scores may not be due to random chance given $\alpha = .05$.

### Diagnostics

Let's test predictive ability. I fit my basic model on all of the data which in turn will be used for prediction.

```{r basic_model_test}
basic_preds <- predict(basic_mod, type = "response") > THRESHOLD
confusionMatrix(admissions$y_admit, reference = basic_preds)
```
